A family of accessibility measures derived from spatial interaction principles

### 2025-02-28 paper/manuscript.qmd
Full draft done - ready for other co-authors review!

```{r load-packages, include=FALSE, cache=FALSE}
library(bibliometrix) # Comprehensive Science Mapping Analysis
library(dplyr) # A Grammar of Data Manipulation
library(flextable) # Functions for Tabular Reporting
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
library(glue) # Interpreted String Literals
library(gt) # Easily Create Presentation-Ready Display Tables
library(here) # A Simpler Way to Find Your Files
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
```

@@

The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations. The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations. Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance. In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of the single proportionality constant $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data. In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.

@@

Building on this, a family of spatial interaction models can be derived. In the framework introduced in @wilson1971, this includes the "unconstrained" (which we refer to as the total flow constrained) model in @eq-phys-gravity-model. The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:

$$
k=\frac{T}{\sum_i\sum_j T_{ij}}
$${#eq-total-flow-balancing-factor}

@@

And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region:
$$
\begin{array}{l}
f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
\end{array}
$$

@@

unc_acc |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Zone") |>
  tab_spanner(label = "{{V[_i^0]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(V_unc_i_1 = md("units: *f_1(c_ij)-weighted physicians"),
             V_unc_i_2 = md("units: *f_2(c_ij)-weighted physicians"),
             V_unc_i_3 = md("units: *f_3(c_ij)-weighted physicians"))  |>
  # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
  #            V_unc_i_2 = "{{f_2 (c_ij )  = 1/c[_ij^2]}}<br>*physicians min^-2*",
  #            V_unc_i_3 = "{{f_3 (c_ij )  = 1/c[_ij^0.1]}}<br>*physicians min^-0.1*",
  #            .fn = md)  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(60),
    everything() ~ px(180)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

@@

Unlike in @eq-access-01, the proportionality constant $k$ is retained and not set arbitrarily to 1:

$$
V^T_{ij} = k^DW_j^{(2)} f(c_{ij})
$${#eq-total-constrained-access}

@@

$$
V^T_i = \sum_j V^T_{ij} = k^D\sum_j W^{(2)}_jf(c_{ij}) = k^DV^0_i
$${#eq-total-constrained-accessibility}

@@

This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework. We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $k$ to yield: 
$$
k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$${#eq-total-opportunity-balancing-factor}

Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974). Hence, our totally-constrained opportunity accessibility model is:

$$
V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
$$

Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term is dimensionless:

$$
\kappa^D_{ij} = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$$

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij}$). The role of $\kappa^D_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region. Constant $\kappa^D_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$. For this reason, we refer to it as a proportional allocation factor.

Referring to our simple numeric example, proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
$$
\begin{array}{l}
k^D = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
k^D = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
}\\
k^D = \frac{490}{0.75} \\
k^D = 653.3256
\end{array}
$$

@@

In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios. Each value is effectively in units of physicians, with the impedance units already accounted for by  $\kappa^D_{ij}$.

@@

One can also directly compare values at a specific zone due to the consistent units. For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case. However, the difference between travel behaviour scenarios differ in direction. Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively. Why? $k^D$ ensures proportional allocation for each travel behaviour scenario. Meaning, while the unconstrained accessibility increases, $k^D$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region). As the decay behaviour decreases, more opportunities are accessible for all zones. In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones. However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).

Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @han1959's accessibility measure. Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.

@@

Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 [@vickermanAccessibilityAttractionPotential1974]. The unconstrained form of market potential has been used in recent research  [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road] to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects. To formulate this case, the total constraint may instead be applied to the mass of the _population_ at $i$ instead of the opportunities at $j$.

$$
M^T_{ij} = k^OW_i^{(1)} f(c_{ij}) = k^OM_{ij}^0
$${#eq-total-constrained-market}

@@

The population-constrained accessibility (market potential) becomes:

$$
M^T_j = \sum_i M^T_{ij} = k^O\sum_i W^{(1)}_if(c_{ij})
$${#eq-total-constrained-market}

@@

Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $k$, we obtain:

$$
k^O=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
$${#eq-total-population-balancing-factor}

@@

$$
M^T_i = k\sum_i W^{(1)}_if(c_{ij}) = \frac{O}{\sum_i\sum_j M^T_{ij}}\sum_i W^{(i)}_i f(c_{ij})
$$

Returning back to the numerical example, the proportionality constant would be solved for each travel behaviour scenario, and the market potential of each zone is expressed as units of population (e.g., the number of people accessible from each zone) in @tbl-simple-example-total-population-accessibility. Readers should not, zonals trends in accessible population and accessible physicians. in the preceding subsection (i.e., calculated using totally-constrained opportunity) are not the same: zone 1, 2, 3 represent destinations and the accessibility values represent the number of accessible people from the vantage of physicians.

<!--maybe we should say a little more? -->

@@

tot_market |>
  select(did, starts_with("M_")) |>
  gt(rowname_col = "did") |>
  tab_stubhead(label = "Destination") |>
  tab_spanner(label = "{{M[_i^D]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(M_tot_i_1 = md("units: *population in 10,000s*"),
             M_tot_i_2 = md("units: *population in 10,000s*"),
             M_tot_i_3 = md("units: *population in 10,000s*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(90),
    everything() ~ px(150)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

@@

### Singly-constrained opportunity: spatial availability(?)

If additional information is introduced into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (@eq-attraction-constraint)- comparable to the single constraint (@eq-constraint2-gravitymodel) from Wilson's framework. This constraint implicitly maintains the total system constraint (@eq-total-constraint-accessibility).

$$
\sum_i V^D_{ij} =  D_j
$${#eq-attraction-constraint}

@@

$$
V^D_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
$${#eq-opportunity-constrained-accessibility}

@@

If we introduce this balancing factor in @eq-opportunity-constrained-accessibility we obtain:

$$
V^D_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
$${#eq-opportunity-constrained-accessibility-with-balancing-factor}

@@

$$
\kappa^D _{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
$${#eq-opportunity-constrained-proportional-allocation-factor}

After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as:

$$
V^D_{i} = \sum_j  \kappa^D_{ij} D_j
$${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}

@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^D_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$. As before, $\kappa^D_{ij}$ is dimensionless and $V_i^D$ is in the units of opportunities $D_j$. The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that there is no competition for them. These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:

$$
v^D_{i} = \frac{V^D_{i}}{W^{(1)}_i}
$$

As an additional point, @eq-attraction-constraint also implies that the total constraint (e.g., $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$) is maintained.

Due to the constraints, $\frac{V_i^D}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system.

@@

opp_acc_ij <- OD |>
  left_join(B_j,
            by = "did") |>
  mutate(V_opp_ij_1 = B_j_1 * O_i * D_j * f1,
         V_opp_ij_2 = B_j_2 * O_i  * D_j * f2,
         V_opp_ij_3 = B_j_3 * O_i  * D_j * f3)

opp_acc <- opp_acc_ij |>
  group_by(oid) |>
  summarize(V_opp_i_1 = sum(V_opp_ij_1),
            V_opp_i_2 = sum(V_opp_ij_2),
            V_opp_i_3 = sum(V_opp_ij_3))

opp_acc  |>
  select(oid, starts_with("V_")) |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>
  tab_spanner(label = "{{V[_i^D]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(V_opp_i_1 = md("units: *physicians*"),
             V_opp_i_2 = md("units: *physicians*"),
             V_opp_i_3 = md("units: *physicians*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(60),
    everything() ~ px(180)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

<!--COMEBACKHERE -->
Similar to the accessibility differences and between pairs of zone using the two different impedance functions: 
```{r}
opp_acc$V_opp_i_1[2] - opp_acc$V_opp_i_1[1]
opp_acc$V_opp_i_2[2] - opp_acc$V_opp_i_2[1]
opp_acc$V_opp_i_3[2] - opp_acc$V_opp_i_3[1]

opp_acc$V_opp_i_1[2] / opp_acc$V_opp_i_1[1]
opp_acc$V_opp_i_2[2] / opp_acc$V_opp_i_2[1]
opp_acc$V_opp_i_3[2] / opp_acc$V_opp_i_3[1]
```

This illustrates how the differences and ratios are now meaningful.

The total accessibility compared to the number of opportunities:
```{r}
opp_acc <- opp_acc |> 
  left_join(LU |>
              select(-D_j),
            by = c("oid" = "id"))

opp_acc  |>
  summarize(V_opp_1 = sum(V_opp_i_1),
            V_opp_2 = sum(V_opp_i_2),
            V_opp_3 = sum(V_opp_i_3))

LU |>
  summarize(O = sum(D_j))
```

This shows how the number of opportunities is preserved.

>>>>

```{r load-packages, include=FALSE, cache=FALSE}
library(bibliometrix) # Comprehensive Science Mapping Analysis
library(dplyr) # A Grammar of Data Manipulation
library(flextable) # Functions for Tabular Reporting
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
library(glue) # Interpreted String Literals
library(gt) # Easily Create Presentation-Ready Display Tables
library(here) # A Simpler Way to Find Your Files
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
```
```{r options}
suppressMessages(library(gt))
options(gt.html_tag_check = FALSE)
```

@@

The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations. The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations. Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance. In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data. In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled: it balances the units.

@@

Building on this, a family of spatial interaction models can be derived. In the framework introduced in @wilson1971, this includes the "unconstrained" (which we refer to as the total flow constrained) model in @eq-phys-gravity-model. The balancing constant $K$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:

$$
K=\frac{T}{\sum_i\sum_j T_{ij}}
$${#eq-total-flow-balancing-factor}

@@

And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios:
$$
\begin{array}{l}
f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
\end{array}
$${#eq-travel-behaviour-scenarios}

@@

unc_acc |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>
  tab_spanner(label = "{{V[_i^0]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}", 
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
             V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
             V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
             .fn = md)  |>
  # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
  #            V_unc_i_2 = "{{f_2 (c_ij )  = 1/c[_ij^2]}}<br>*physicians min^-2*",
  #            V_unc_i_3 = "{{f_3 (c_ij )  = 1/c[_ij^0.1]}}<br>*physicians min^-0.1*",
  #            .fn = md)  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(60),
    everything() ~ px(180)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

@@

Unlike in @eq-access-01, the proportionality constant $k$ is retained, not set arbitrarily to 1, and represented as $K^T$:

$$
V^T_{ij} = K^T \cdot W_j^{(2)} \cdot f(c_{ij})
$${#eq-total-constrained-access}

@@

$$
V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T \cdot V^0_i
$${#eq-total-constrained-accessibility}

@@

This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework. We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $K^T$ to yield: 
$$
K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$${#eq-total-opportunity-balancing-factor}

Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974). Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our totally-constrained opportunity accessibility model is:

$$
V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
$$

Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term, the proportional allocation factor for the totally-constrained opportunity case $\kappa^T$ is dimensionless:

$$
\kappa^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$$

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$). The role of $\kappa^T$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region. Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$. For this reason, we refer to it as a proportional allocation factor.

Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be:
$$
\begin{array}{l}
K^T = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
K^T = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
}\\
K^T = \frac{490}{0.6233422} \\
K^T = 786.085
\end{array}
$$

@@

In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios. Each value is effectively in units of physicians, with the impedance units already accounted for by  $\kappa^T$.

@@

One can also directly compare values at a specific zone due to the consistent units. For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case. However, the difference between travel behaviour scenarios differ in direction. Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively. Why? $\kappa^T$ ensures proportional allocation for each travel behaviour scenario. Meaning, while the unconstrained accessibility increases, $\kappa^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region). As the decay behaviour decreases, more opportunities are accessible for all zones. In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones. However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).

Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @hansen1959's accessibility measure. Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.

@@

Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 [@vickermanAccessibilityAttractionPotential1974]. The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$,  has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. To formulate this case, the total constraint may instead be applied to the mass of the _population_ at $i$ instead of the opportunities at $j$.

$$
M^T_{ij} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ij}^0
$${#eq-total-constrained-market}

\noindent with $M_j^0$ being the $i$ $j$ transpose of $V^0_{ij}$:

@@

The totally-constrained population accessibility (market potential) becomes:

$$
M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
$${#eq-total-constrained-market}

@@

Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $\hat K^T$, we obtain:

$$
\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
$${#eq-total-population-balancing-factor}

@@

$$
M^T_i = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
$$
Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless:
$$
\hat \kappa^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
$$

Returning back to the numerical example, the proportionality constant would be solved for each travel behaviour scenario, and the market potential of each zone is expressed as units of population (e.g., the number of people accessible from each zone) in @tbl-simple-example-total-population-accessibility. Readers may note the difference in trends in accessible population (@tbl-simple-example-total-population-accessibility) and accessible physicians (i.e., the preceding subsection, @tbl-simple-example-total-opportunity-accessibility). In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians. Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios. However, similar to the totally-constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons. As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.

@@

tot_market |>
  select(did, starts_with("M_")) |>
  gt(rowname_col = "did") |>
  tab_stubhead(label = "Destination") |>
  tab_spanner(label = "{{M[_i^S]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(M_tot_i_1 = md("units: *population in 10,000s*"),
             M_tot_i_2 = md("units: *population in 10,000s*"),
             M_tot_i_3 = md("units: *population in 10,000s*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(90),
    everything() ~ px(150)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

@@

### Singly-constrained opportunity: A.K.A spatial availability

If additional information is introduced into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (@eq-opportunity-constraint)- comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.

$$
\sum_i V^S_{ij} =  D_j
$${#eq-opportunity-constraint}

@@

$$
V^S_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
$${#eq-opportunity-constrained-accessibility}

@@

Introducing the balancing factor in @eq-opportunity-constrained-accessibility, we obtain:

$$
V^S_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
$${#eq-opportunity-constrained-accessibility-with-balancing-factor}

@@

$$
\kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
$${#eq-opportunity-constrained-proportional-allocation-factor}

After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as an origin summary expression of proportionally allocated known opportunities (i.e., $D_j$):

$$
V^S_{i} = \kappa^S_{i} \sum_j   D_j
$${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}

@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{i}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$. As in the totally=constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$. The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered. These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:

$$
v^S_{i} = \frac{V^S_{i}}{W^{(1)}_i}
$$

To reinforce this point, @eq-opportunity-constraint also implies that the total system constraint (e.g., $\sum_i V^S_{i} = \sum_i\sum_j  V^S_{ij} = D$ in @eq-total-constraint-accessibility) is maintained.

Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system.

@@

opp_acc_ij <- OD |>
  left_join(B_j,
            by = "did") |>
  mutate(k_opp_ij_1 = B_j_1 * O_i * f1 /3,
         k_opp_ij_2 = B_j_2 * O_i  * f2 /3,
         k_opp_ij_3 = B_j_3 * O_i  * f3 /3,
         V_opp_ij_1 = B_j_1 * O_i * D_j * f1,
         V_opp_ij_2 = B_j_2 * O_i  * D_j * f2,
         V_opp_ij_3 = B_j_3 * O_i  * D_j * f3)

opp_acc <- opp_acc_ij |>
  group_by(oid) |>
  summarize(O_i = first(O_i) |> as.character(), 
            V_opp_i_1 = sum(V_opp_ij_1),
            V_opp_i_2 = sum(V_opp_ij_2),
            V_opp_i_3 = sum(V_opp_ij_3), 
            .groups = "drop")

opp_acc_with_ks <-  opp_acc_ij |>
  group_by(oid) |>
  summarize(O_i = first(O_i) |> as.character(), 
            k_opp_ij_1 = sum(k_opp_ij_1),
            k_opp_ij_2 = sum(k_opp_ij_2),
            k_opp_ij_3 = sum(k_opp_ij_3))

opp_acc |>
  #select(oid, O_i, starts_with("V_")) |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>

  tab_spanner(label = "{{V[_i^S]}}",
              columns = 3:5,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 4,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 5,
              level = 1) |>
  fmt_number(decimals = 3) |>

  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
             V_opp_i_1 = md("units: *physicians*"),
             V_opp_i_2 = md("units: *physicians*"),
             V_opp_i_3 = md("units: *physicians*"),
             )  |>
  cols_align(align = "center",
             columns = 2:5) |>
  cols_width(
    stub() ~ px(60*0.8),
    everything() ~ px(180*0.8)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

Imposing the single constraint $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values like previously discussed in the case of the totally-constrained opportunity case. Namely, when the single constraint is applied to the opportunities at $j$, each resulting accessibility value at $i$ is in units of physicians, with the impedance units already accounted for.
